---
title: "Xenium post-processing"
author:  "Shubhada Kulkarni"
date: "23 July 2024"
format: html
editor: visual
---

## INFO

This Quarto computational documents includes post-processing of already clustered Xenium data generated using `process_Xenium_data.R` script. This post-processing includes

* [] Merging and batch-effect correction of data per slide.
* [] Post-merging processing
* [] Cell-type annotations

## Cell-type annotations
Loading required libraries 
```{r}
library(ggplot2, lib.loc = "/beegfs/homes/skulkarni/R/x86_64-pc-linux-gnu-library/4.3")
library(Seurat, lib.loc = "/beegfs/homes/skulkarni/R/x86_64-pc-linux-gnu-library/4.3")
library(patchwork)
library(markdown)
library(rmarkdown)
library(dplyr, lib.loc = "/beegfs/homes/skulkarni/R/x86_64-pc-linux-gnu-library/4.3")
library(presto, lib.loc = "/beegfs/homes/skulkarni/R/x86_64-pc-linux-gnu-library/4.3")
```

### Slide 0027119 (7/3 male R231Q/wt)
Read in stored Xenium objects for both regions for _0027119_
```{r eval=F}
region1 <- readRDS("/prj/XeniumProbeDesign/kidney_Nphs2-mice_Xenium_Martin/xenium_objects_R/output-XETG00046__0027119__Region_1__20240621__120943_processed.rds")
region2 <- readRDS("/prj/XeniumProbeDesign/kidney_Nphs2-mice_Xenium_Martin/xenium_objects_R/output-XETG00046__0027119__Region_2__20240621__120943_processed.rds")
```

Dimplots of both regions
```{r fig.align='center', eval=FALSE}
p1 <- DimPlot(region1) + labs(title = "Region 1")
p2 <- DimPlot(region1) + labs(title = "Region 2")
p1 + p2
```
As DimPlots of region1 and region2 are not so different + they seem to share many marker genes, the cell-type annotations will be performed on 1 region and will be used for the second region as well.

#### Merging the two regions and post-processing
A separate R script was used to perform merging of two regions, as memory was running out. This R script was submitted as a SLURM job. 
Output of this script is a processed merged Xenium objects that contain both regions.
Following operations have been performed during above processing:
* Merging
* Normalising and scaling data
* Re-computing PCA and UMAP
* Finding neighbors
* Finding clusters

In next steps, cluster annotation will be performed.
#### Marker gene analysis
Read in the RDS merged object created by above R script
```{r eval=FALSE}
xenium.obj.0027119 <- readRDS("/prj/XeniumProbeDesign/kidney_Nphs2-mice_Xenium_Martin/xenium_objects_R/0027119_merged_processed.rds")
```

Finding marker genes using presto package (much faster than Seurat functions)
```{r eval=FALSE}
all_markers_0027119 <- presto::wilcoxauc(xenium.obj.0027119,  seurat_assay = "SCT", group_by = "seurat_clusters")
write.csv(all_markers_0027119, "/prj/XeniumProbeDesign/kidney_Nphs2-mice_Xenium_Martin/xenium_objects_R/0027119_merged_markergenes.csv", quote = F)
```

#### Cell-type annotations
For selection of marker genes per cell-type, I looked into multiple papers publishing scRNA-seq mouse kidney. However, after trying multiple ones, following paper had the most overlapping cell-types https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10690238/#SD1 

For the sake of saving time to load featureplot for every gene, only one gene plot is shown

Function to plot expression levels on UMAP and spatial level
```{r eval=FALSE}
plot_featureplots <- function(gene){
  p1 <- FeaturePlot(xenium.obj.0027119, c(gene), label = T) + labs(title = "On UMAP Level")
  p2 <- ImageFeaturePlot(xenium.obj.0027119, features = gene, dark.background = F) + labs(title = "On Spatial Level")
  return(p1 | p2)
}
```

Podocytes markers -> Nphs2, Ddn, Clic3 and Rab3b
```{r eval=FALSE}
plot_featureplots("Nphs2")
```
Cluster 9 and 14 are podocytes.

PTS1/2/3 – proximal tubule segment
1 -> Slc22a8, Lrp2, Cyp4b1
2 -> same
3 -> Aadat, Aqp1
```{r eval=FALSE}
plot_featureplots("Slc22a8")
```

```{r eval=FALSE}
plot_featureplots("Aqp1")
```
Cluster 0, 20, 28, 2, 21, 15, 26, 36, 11, 30, 31, 24, 35, 1, 34 are proximal tubule segment clusters 1, 2and 3.

thin limb of the loop of Henle (LOH) associated cells -> Bst1, Aqp1, Cryab,  Pax8, Upk3b
```{r eval=FALSE}
plot_featureplots("Bst1")
```
Cluster 17 is LOH cells

Thick ascending limb of the loop of Henle (TAL) 
TAL1 -> Sostdc1, Gpx6, Ppargc1a, Prox1
TAL2 -> ""Slc5a1"","Dusp15" + TAL1
TAL3 -> "Gpx6","Scin"
```{r eval=FALSE}
plot_featureplots("Ppargc1a")
```
Clusters 4, 8, 10, 25, 33 are TAL clusters

Distal convoluted and connecting tubule (DCT-CNT) -> Calb1, Hsd11b2, Ldhb
```{r eval=FALSE}
plot_featureplots("Hsd11b2")
```
Clusters 5 and 29 are DCT-CNT cells

Principal cells (PC) -> "Fxyd4", "Hsd11b2", "Aqp3" 
```{r eval=FALSE}
plot_featureplots("Fxyd4")
```

Intercalated cells (IC) -> "Slc4a1", "Car2"
```{r eval=FALSE}
plot_featureplots("Slc4a1")
```
Both IC and PC cells overlap with DCT-CNT cells. So clusters 5 and 29 are DCT, IC and PC cells.

Urothelium -> Upk1b, Upk3b
```{r eval=FALSE}
plot_featureplots("Upk1b")
```
Cluster 22 is urothelium cells

Fibroblasts -> Fbln5, Angptl2, "Mylk", "Col4a2"
```{r eval=FALSE}
plot_featureplots("Fbln5")
```
Clusters 6, 12, 13 and 16 are fibroblasts

Pericytes -> Myh11, Ren1, Gja5
```{r eval=FALSE}
plot_featureplots("Fbln5")
```
Cluster 8 is Pericytes

Vascular cells -> Plvap, Kdr
```{r eval=FALSE}
plot_featureplots("Fbln5")
```
Cluster 3 is vascular cells

Macrophages -> "C5ar1", "Cybb", "Mpeg1"
```{r eval=FALSE}
plot_featureplots("Cybb")
```
Cluster 7 is macrophages

Dendritic cell genes were not present in the gene panel so not found.
T cells were also not found in this dataset.

Rename all clusters.
```{r eval=FALSE}
new.cluster.ids <- c("PT", "PT", "PT", "Vascular", "TAL", "DCT-IC-PC",  "Fibroblasts", "Macrophages", "TAL", 
                     "Podocytes", "TAL", "PT", "Fibroblasts", "Fibroblasts", "Podocytes", "PT", "Fibroblasts",
                     "LOH", "Unidentified1", "Injured cells", "PT", "PT", "Urothelium", "Unidentified2", "PT", "TAL", "PT", "Unidentified3", "PT",
                     "DCT-IC-PC",  "PT", "PT", "PT", "TAL", "PT", "PT", "PT")
names(new.cluster.ids) <- levels(xenium.obj.0027119)
xenium.obj.0027119 <- RenameIdents(xenium.obj.0027119, new.cluster.ids)
```

DimPlot after changing cluster names
```{r fig.align='center'}
DimPlot(xenium.obj.0027119, label = T)
```

ImagePlot after changing cluster names
```{r fig.align='center'}
ImageDimPlot(xenium.obj.0027119, dark.background = F)
```


<!-- ### Slide 0027120 (4/1 male compound het) -->

<!-- Read in stored Xenium objects for both regions for _0027120_ -->
<!-- ```{r eval=F} -->
<!-- region1 <- readRDS("/prj/XeniumProbeDesign/kidney_Nphs2-mice_Xenium_Martin/xenium_objects_R/output-XETG00046__0027120__Region_1__20240621__120943_processed.rds") -->
<!-- region2 <- readRDS("/prj/XeniumProbeDesign/kidney_Nphs2-mice_Xenium_Martin/xenium_objects_R/output-XETG00046__0027120__Region_2__20240621__120943_processed.rds") -->
<!-- ``` -->

<!-- Dimplots of both regions -->
<!-- ```{r fig.align='center', eval=FALSE} -->
<!-- p1 <- DimPlot(region1) + labs(title = "Region 1") -->
<!-- p2 <- DimPlot(region1) + labs(title = "Region 2") -->
<!-- p1 + p2 -->
<!-- ``` -->

<!-- Merging -->
<!-- ```{r} -->
<!-- xenium_0027120 <- merge(region1, y = region2, add.cell.ids = c("region1", "region2"), project = "Xenium_0027120") -->
<!-- ``` -->

